Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2022-02-09 referee-response.Rmd

14. p11, line 31. "... SSC on the spectral embedding of $A$". Could you support this assertion by references?  
\textcolor{blue}{
There are no references for this, as it is, to the best of our knowledge, a novel idea. 
We emphasize that while SSC has been proposed for community detection
for PABM (see e.g., \citep{noroozi2019estimation}), these work apply SSC
directly on the rows of $A$ (which are binary vectors in
$\mathbb{R}^{n}$) rather than on the ASE of $A$ (which are real-valued
vectors in $\mathbb{R}^{d}$ where $d \ll n$). The motivation behind
our use of SSC comes  from Theorem 1 wherein we show that the latent
vectors for the PABM lie on the union of orthogonal subspaces in $\mathbb{R}^{d}$ and hence,
by the consistency properties of ASE, the rows of the estimated
eigenvectors $\hat{V}$ also lie close to the union of orthogonal
subspaces in $\mathbb{R}^{d}$. In contrast, while the rows of $P$ do lie exactly on the
union of $K$ subspaces in $\mathbb{R}^{n}$, the rows of $A$
(which are now noisy binary vectors) could be quite far from the
corresponding rows of $P$ and hence need not lie close to the union
of any $K$ subspaces in $\mathbb{R}^{n}$ 
}

15. p12, line 13. It could help recall what the $z_i, z_j$ are.  
\textcolor{blue}{
We have changed $z_i$ to ``community labels''.
}

16. p14. Do you think that the OSC algorithm can retrieve other kinds of structure than assortative ones?  
\textcolor{blue}{
The OSC algorithm works for any PABM parameters setting and is thus
agnostic to the particular structure (such as assortativity vs
disassortativity). We have revised the manuscript to include an
additional simulation demonstrating this claim (see Section 4.3). In
particular Section~4.3 considers a PABM setting wherein
$\lambda_{kk}$ is a vector in $\mathbb{R}^{n_k}$ whose elements
are i.i.d. $\mathrm{Beta}(1,2)$ random variables and $\lambda_{k\ell}$
(for $k \not = \ell$) is a
vector in $\mathbb{R}^{n_k}$ whose elements are i.i.d. $\mathrm{Beta}(2,1)$
random variables. This result in an average within-community popularity 
parameter of $1/3$ and a between-community popularity
parameter of $2/3$. The results for this simulation setting are
very similar to those presented in the original manuscript and is also
consistent with the theoretical results in the paper.
}  
\textcolor{blue}{
On a related note, it is our view that the distinction between
assortative and disassortative is not always meaningful for the
PABM. Indeed, unlike the SBM, in the PABM each vertex is free to have a higher affinity to its
own community or to other communities. For example, suppose vertex $i$
belongs to community $1$. Then the average probability of an edge
between vertex $i$ and another vertex in community $1$ is $n^{-1}
\rho_n 
\lambda_{i1} \sum_{j} \lambda_{j1}$ while the average probability of an edge between vertex $i$ and another vertex in community $k \not =
1$ is $n^{-1} \rho_n \lambda_{ik} \sum_{j} \lambda_{j1}$ and thus, on
average, vertex $i$ has more affinity with vertices in community $k \not =
1$ whenever $\lambda_{i1} < \lambda_{ik}$
}

@@

17. Sections 4 and 5. Why do you use either the ARI criterion or the misclassification rate to assess the recovery of the nodes clustering?  
\textcolor{blue}{
The first two datasets that we analyzed in section 5 had also been analyzed
in previous papers on the PABM. In particular the
Leeds Butterfly dataset was analyzed in \citep{noroozi2019estimation}
and while \citep{noroozi2019estimation} provides enough details
for us to preprocess the data to match their analysis, they do not
provide code to run their clustering algorithm (which is based on SSC
applied to the rows of $A$). Therefore we used their reported
benchmark values, which are given in terms of ARI. 
In order to make an apples-to-apples comparison, we also report the ARI for OSC and SSC-ASE here as well. 
For all other analyses (simulation and real data), we use the
misclassification error rate/count. 
}

18. Do you consider sparsity in the simulated adjacency matrices?  
\textcolor{blue}{
We have added another simulation in the supplemental materials. 
In this simulation, we fix $n = 2048$ and $K = 3$ and vary $\rho \in
(0, 1)$. In particular we see that the performance of all algorithms
deteriorate significantly as the graphs become sparser. Nevertheless
it appears that the accuracy of OSC is always comparable and/or better
than the remaining algorithms for all levels of the sparsity
considered.
}

19. Why do you limit your simulation to assortative structure? 
Is it possible to see the results with a larger value of $K$? 
To what extent does the value of $K$ impact the computational burden for the algorithms?  
\textcolor{blue}{
As we discussed in the response to comment
$16$, we have added a third simulation to address the issue of disassortativity (see Section
4.3 of the revised manuscript). The performance of all algorithms
for the setting in Section~4.3 is qualitatively similar to those
presented earlier in Section~4.1 and Section~4.2 and is consistent with the theory. 
}  
\textcolor{blue}{
For the simulation studies, we only considered $K \leq 4$ as larger $K$ will, 
in general, require larger values of $n$ to achieve comparable estimation accuracy as lower values of $K$. 
Indeed, consider for example estimating the entries of the edge probabilities matrix $P$. 
Then the rank of $P$ is equal to $K^2$ and hence, for a given $n$, estimation of $P$ when $K = 4$ is considerably simpler than estimation of $P$ when $K = 6$. 
We have performed some cursory numerical experiments that suggest that OSC does behave similarly to the $K \leq 4$ cases, given sufficient $n$, which is consistent with the theory. 
}  
\textcolor{blue}{
As for the computational complexity of the algorithms, 
the main bottleneck for OSC and
SSC-ASE is the construction of the affinity matrix $B$. 
For OSC, this involves a (truncated) spectral decomposition to extract the
$K^2$ largest eigenvalues in modulus (and their corresponding
eigenvectors) followed by a matrix multiplication of $n \times K^2$
and $K^2 \times n$ matrices, so the time complexity is of the order $O(n^2 K^2)$. 
For SSC-ASE (applying SSC on the rows of the adjacency spectral
embedding $\hat{V}$) we again start with spectral decomposition, followed by
$n$ LASSO problems with a design matrix of size 
$K^2 \times (n - 1)$, so the complexity is $O(n^2 K^2 + n^2 K^4 + n K^6)$ \cite{10.1214/009053604000000067}. 
SSC-A (applying SSC on the rows of $A$) as proposed in \cite{noroozi2019estimation} involves solving $n$ LASSO regression problems each with design matrices of size $n \times (n - 1)$, so the complexity here is $O(n^4)$. 
In practice the choice of parameter $K$ does not affect
runtimes too much in our simulation studies, 
largely because SSC-A takes up the bulk of the runtime. 
}

### 2022-02-09 referee-response.tex

  additional simulation demonstrating this claim (see Section 4.3). In
  particular Section~4.3 considers a PABM setting wherein
  $\lambda_{kk}$ is a vector in $\mathbb{R}^{n_k}$ whose elements
  are i.i.d. $\mathrm{Beta}(1,2)$ random variables and $\lambda_{k\ell}$

@@

  (for $k \not = \ell$) is a
  vector in $\mathbb{R}^{n_k}$ whose elements are i.i.d. $\mathrm{Beta}(2,1)$
  random variables. This result in an average within-community popularity 
  parameter of $1/3$ and a between-community popularity
  parameter of $2/3$. The results for this simulation setting are

@@

  algorithms?\\
  \textcolor{blue}{
  As we discussed in the response to comment
  $16$, we have added a third simulation to address the issue of disassortativity (see Section

@@

  presented earlier in Section~4.1 and Section~4.2 and is consistent with the theory. 
  }\\
  \textcolor{blue}{
  For the simulation studies, we only considered $K \leq 4$ as larger $K$ will, 
  in general, require larger values of $n$ to achieve comparable estimation accuracy as lower values of $K$. 
  Indeed, consider for example estimating the entries of the edge probabilities matrix $P$. 
  Then the rank of $P$ is equal to $K^2$ and hence, for a given $n$, estimation of $P$ when $K = 4$ is considerably simpler than estimation of $P$ when $K = 6$. 
  We have performed some cursory numerical experiments that suggest that OSC does behave similarly to the $K \leq 4$ cases, given sufficient $n$, which is consistent with the theory. 
  As for the computational complexity of the algorithms, 
  the main bottleneck for OSC and

@@

  $K^2 \times (n - 1)$, so the complexity is $O(n^2 K^2 + n^2 K^4 + n K^6)$ \cite{10.1214/009053604000000067}. 
  SSC-A (applying SSC on the rows of $A$) as proposed in \cite{noroozi2019estimation} involves solving $n$ LASSO regression problems each with design matrices of size $n \times (n - 1)$, so the complexity here is $O(n^4)$. 
  In practice the choice of parameter $K$ does not affect
  runtimes too much in our simulation studies, 
  largely because SSC-A takes up the bulk of the runtime. 

@@

  right value for the hyperparameter $\vartheta$ for SSC-ASE. 
  }\\
  \textcolor{blue}{
  More specifically, under ideal circumstances, we can set the hyperparameter $\vartheta$ to
  the theoretical values that ensure, with high probability, the rows of $\hat{V}$ satisfy the subspace

@@

  clustering the rows of this embedding using Gaussian mixture models. 
  }\\
  \textcolor{blue}{
  Finally we emphasize that in our simulations the proportion of mislabeled
  vertices does decreases as $n$ increases. For instance, in section 4.1, if $K = 2$ then the average number of mis-clustered vertices for SSC-ASE is around 20 when $n = 128$ and around 100 when
  $n = 4096$; this correspond to an error rate of around $16\%$ and
  $2\%$, respectively. In other words SSC-ASE also works quite well for $K =
  2$ as the proportion of mis-clustered vertices decreases to $0$ as $n$
  increases; it is only when we compared the number of mis-clustered
  vertices that $K = 2$ performs worse compared to $K \geq 3$. 

### 2022-02-11 referee-response.Rmd

17. Sections 4 and 5. Why do you use either the ARI criterion or the misclassification rate to assess the recovery of the nodes clustering?  
\textcolor{blue}{
The first two datasets that we analyzed in section 5 had also been analyzed
in previous papers on the PABM. In particular the
Leeds Butterfly dataset was analyzed in \citep{noroozi2019estimation}
and while \citep{noroozi2019estimation} provides enough details
for us to preprocess the data to match their analysis, they do not
provide code to run their clustering algorithm (which is based on SSC
applied to the rows of $A$). Therefore we used their reported
benchmark values, which are given in terms of ARI. 
In order to make an apples-to-apples comparison, we also report the ARI for OSC and SSC-ASE here as well. 
For all other analyses (simulation and real data), we use the
misclassification error rate/count. 
}

18. Do you consider sparsity in the simulated adjacency matrices?  
\textcolor{blue}{
We have added another simulation in the supplemental materials. 
In this simulation, we fix $n = 2048$ and $K = 3$ and vary $\rho \in
(0, 1)$. In particular we see that the performance of all algorithms
deteriorate significantly as the graphs become sparser. Nevertheless
the accuracy of OSC is always comparable and/or better than the remaining algorithms for all levels of the sparsity
considered.
}

19. Why do you limit your simulation to assortative structure? 
Is it possible to see the results with a larger value of $K$? 
To what extent does the value of $K$ impact the computational burden for the algorithms?  
\textcolor{blue}{
As we discussed in the response to comment
$16$, we have added a third simulation to address the issue of disassortativity (see Section
4.3 of the revised manuscript). The performance of all algorithms
for the setting in Section~4.3 is qualitatively similar to those
presented earlier in Section~4.1 and Section~4.2 and is consistent with the theory. 
}  
\textcolor{blue}{
As for the computational complexity of the algorithms, the main bottleneck for OSC and
SSC-ASE is the construction of the affinity matrix $B$. 
For OSC, this involves a (truncated) spectral decomposition to extract the
$K^2$ largest eigenvalues in modulus (and their corresponding
eigenvectors) followed by a matrix multiplication of $n \times K^2$
and $K^2 \times n$ matrices, so the time complexity is of the order $O(n^2 K^2)$. 
For SSC-ASE (applying SSC on the rows of the adjacency spectral
embedding $\hat{V}$) we again start with spectral decomposition, followed by
$n$ LASSO problems with a design matrix of size 
$K^2 \times (n - 1)$, so the complexity is $O(n^2 K^2 + n^2 K^4 + n K^6)$ \cite{10.1214/009053604000000067}. 
SSC-A (applying SSC on the rows of $A$) as proposed in \cite{noroozi2019estimation} involves solving $n$ LASSO regression problems each with design matrices of size $n \times (n - 1)$, so the complexity here is $O(n^4)$. 
In practice the choice of parameter $K$ does not affect
runtimes too much. For the simulation studies we only considered $K
\leq 4$ as values of $K \geq 5$ will in general, require larger values of $n$ to achieve 
comparable estimation accuracy as $K \leq 4$.  Indeed, consider for
example estimating the entries of the edge probabilities matrix
$P$. Then the rank of $P$ is equal to $K^2$ and hence, for a given $n$, estimation of
$P$ when $K = 4$ is considerably simpler than estimation of $P$ when
$K = 8$. Cursory numerical experiments suggest that OSC and
SSC-ASE behave similarly for larger $K$ provided that $n$ is large enough. 
}

### 2022-02-11 referee-response.Rmd

17. Sections 4 and 5. Why do you use either the ARI criterion or the misclassification rate to assess the recovery of the nodes clustering?  
\textcolor{blue}{
The first two datasets that we analyzed in section 5 had also been analyzed
in previous papers on the PABM. In particular the
Leeds Butterfly dataset was analyzed in \citep{noroozi2019estimation}
and while \citep{noroozi2019estimation} provides enough details
for us to preprocess the data to match their analysis, they do not
provide code to run their clustering algorithm (which is based on SSC
applied to the rows of $A$). Therefore we used their reported
benchmark values, which are given in terms of ARI. 
In order to make an apples-to-apples comparison, we also report the ARI for OSC and SSC-ASE here as well. 
For all other analyses (simulation and real data), we use the
misclassification error rate/count. 
}

18. Do you consider sparsity in the simulated adjacency matrices?  
\textcolor{blue}{
We have added another simulation in the supplemental materials. 
In this simulation, we fix $n = 2048$ and $K = 3$ and vary $\rho \in
(0, 1)$. In particular we see that the performance of all algorithms
deteriorate significantly as the graphs become sparser. Nevertheless
the accuracy of OSC is always comparable and/or better than the remaining algorithms for all levels of the sparsity
considered.
}

19. Why do you limit your simulation to assortative structure? 
Is it possible to see the results with a larger value of $K$? 
To what extent does the value of $K$ impact the computational burden for the algorithms?  
\textcolor{blue}{
As we discussed in the response to comment
$16$, we have added a third simulation to address the issue of disassortativity (see Section
4.3 of the revised manuscript). The performance of all algorithms
for the setting in Section~4.3 is qualitatively similar to those
presented earlier in Section~4.1 and Section~4.2 and is consistent with the theory. 
}  
\textcolor{blue}{
For the simulation studies, we only considered $K \leq 4$ as larger $K$ will, 
in general, require larger values of $n$ to achieve comparable estimation accuracy as lower values of $K$ and SSC-A 
(where SSC is applied to the rows of the adjacency matrix) takes a very long time to run when $n$ is large  (see the discussion about computational complexity below). 
Our theoretical results are limit results and work for any choice of $K$ as long as $n$ is large. The reason for needing large $n$ when $K$ increases is that the rank of $P$ is equal to $K^2$ and hence, for a given $n$, estimation of $P$ when $K$ is say $4$ is considerably simpler than estimation of $P$ when $K$ is say $8$. 
We have performed some cursory numerical experiments and found that OSC and SSC-ASE for $K = 8$ and $n = 4096$ behave similarly to $K \leq 4$ and $n = 1024$.
}  
\textcolor{blue}{
As for the computational complexity of the algorithms, 
the main bottleneck for OSC and
SSC-ASE is the construction of the affinity matrix $B$. 
For OSC, this involves a (truncated) spectral decomposition to extract the
$K^2$ largest eigenvalues in modulus (and their corresponding
eigenvectors) followed by a matrix multiplication of $n \times K^2$
and $K^2 \times n$ matrices, so the time complexity is of the order $O(n^2 K^2)$. 
For SSC-ASE (applying SSC on the rows of the adjacency spectral
embedding $\hat{V}$) we again start with spectral decomposition, followed by
$n$ LASSO problems with a design matrix of size 
$K^2 \times (n - 1)$, so the complexity is $O(n^2 K^2 + n^2 K^4 + n K^6)$ \cite{10.1214/009053604000000067}. 
SSC-A (applying SSC on the rows of $A$) as proposed in \cite{noroozi2019estimation} involves solving $n$ LASSO regression problems each with design matrices of size $n \times (n - 1)$, so the complexity here is $O(n^4)$. 
In practice the choice of parameter $K$ does not affect runtimes too much in our simulation studies, 
largely because SSC-A takes up the bulk of the runtime. 
}

### 2022-02-11 referee-response.Rmd

# Referee 1 Comments

## Main comments

The interest of this paper is to provide new perspectives on PABM
(theoretical and algorithmic). The theoretical results are
sound. However, the algorithmic results shown in the simulations are
not really impressive since they are done in settings that do not
appear very challenging. Indeed, they are done with fixed numbers of
communities which are quite low (no more than 4) and only with
assortative structure (larger probabilities of connections within a
community than between two communities). Therefore, I think the paper
needs a more challenging simulation study and the question of the
choice of the number of communities needs to be addressed at least
with some heuristic that works in simulation studies if theoretical
results are not reachable. Moreover, some points on the exposition
could be made clearer. I have specific comments below where I ask for
some clarifications and detail how the simulation study could be
improved in my opinion.

\textcolor{blue}{
We thank the referee for the feedback. The referee's comments are
answered below. 
In particular, we made considerable effort to address the referee's concerns about the simulation studies. 
}

@@

17. Sections 4 and 5. Why do you use either the ARI criterion or the misclassification rate to assess the recovery of the nodes clustering?  
\textcolor{blue}{
The first two datasets that we analyzed in section 5 had also been analyzed
in previous papers on the PABM. In particular the
Leeds Butterfly dataset was analyzed in \citep{noroozi2019estimation}
and while \citep{noroozi2019estimation} provides enough details
for us to preprocess the data to match their analysis, they do not
provide code to run their clustering algorithm (which is based on SSC
applied to the rows of $A$). Therefore we used their reported
benchmark values, which are given in terms of ARI. 
In order to make an apples-to-apples comparison, we also report the ARI for OSC and SSC-ASE here as well. 
For all other analyses (simulation and real data), we use the
misclassification error rate/count. 
}

18. Do you consider sparsity in the simulated adjacency matrices?  
\textcolor{blue}{
We have added another simulation in the supplemental materials. 
In this simulation, we fix $n = 2048$ and $K = 3$ and vary $\rho \in
(0, 1)$. In particular we see that the performance of all algorithms
deteriorate significantly as the graphs become sparser. Nevertheless
the accuracy of OSC is always comparable and/or better than the remaining algorithms for all levels of the sparsity
considered.
}

19. Why do you limit your simulation to assortative structure? 
Is it possible to see the results with a larger value of $K$? 
To what extent does the value of $K$ impact the computational burden for the algorithms?  
\textcolor{blue}{
As we discussed in the response to comment
$16$, we have added a third simulation to address the issue of disassortativity (see Section
4.3 of the revised manuscript). The performance of all algorithms
for the setting in Section~4.3 is qualitatively similar to those
presented earlier in Section~4.1 and Section~4.2 and is consistent with the theory. 
}  
\textcolor{blue}{
For the simulation studies, we only considered $K \leq 4$ as larger
$K$ will, in general, require larger values of $n$ to achieve
comparable estimation accuracy as lower values of $K$. The running
time for a simulation studies when $n$ is large then becomes quite
prohibitive, and this is due mainly to the running-time of the SSC-A
algorithm (see the discussion about computationaly
complexity); the running time of the SSC-ASE and OSC is still
reasonable when we increases $n$. We believe that our current simulation
studies (including the new section 4.3) is sufficiently
comprehensive to illustrate our theoretical results results; indeed
our theoretical results are large-sample results and is valid for any
choice of $K$. The only reason for needing large
$n$ when $K$ increases is that the rank of $P$ is equal to $K^2$ and
hence, for a given $n$, estimation of $P$ when $K$ is say $4$ is
considerably simpler than estimation of $P$ when $K$ is say $8$. 
}  
\textcolor{blue}{
For the computational complexity of the algorithms, 
the main bottleneck for OSC and
SSC-ASE is the construction of the affinity matrix $B$. 
For OSC, this involves a (truncated) spectral decomposition to extract the
$K^2$ largest eigenvalues in modulus (and their corresponding
eigenvectors) followed by a matrix multiplication of $n \times K^2$
and $K^2 \times n$ matrices, so the time complexity is of the order $O(n^2 K^2)$. 
For SSC-ASE (applying SSC on the rows of the adjacency spectral
embedding $\hat{V}$) we again start with spectral decomposition, followed by
$n$ LASSO problems with a design matrix of size 
$K^2 \times (n - 1)$, so the complexity is $O(n^2 K^2 + n^2 K^4 + n K^6)$ \cite{10.1214/009053604000000067}. 
SSC-A (applying SSC on the rows of $A$) as proposed in \cite{noroozi2019estimation} involves solving $n$ LASSO regression problems each with design matrices of size $n \times (n - 1)$, so the complexity here is $O(n^4)$. 
In practice the choice of parameter $K$ does not affect runtimes too much in our simulation studies, 
largely because SSC-A takes up the bulk of the runtime. 
}

@@

of $A$. 
}

* This is clearly beyond the scope of the paper, but I wonder if there is a straightforward way to adapt the proposed methods to directed models or weighted graphs or any other types of graphs than binary undirected ones?  
\textcolor{blue}{
For directed graphs, the ASE provides a consistent estimator and thus should also work here \citep{doi:10.1080/01621459.2012.699795}. 
For weighted graphs, if $A_{ij}$ can be an estimator for $P_{ij}$ (e.g., if $A_{ij} \sim \mathrm{Poisson}(P_{ij})$), then all of the theory should transfer over to those models as well, 
and we believe that the theory can be adapted to other models where $g(A_{ij})$ is an estimator for $P_{ij}$. 
We hope to address these topics in future research. 
}

* In the simulation study, SSC-ASE works well for $K \geq 3$, but not for $K = 2$. 
The authors explain the bad performance by the use of gaussian mixtures in the last clustering step. 
However, I don't understand why this has an impact on the performance for $K = 2$, but not for $K \geq 3$.  
\textcolor{blue}{
We are not entirely clear on why this happens for $K = 2$, but when
comparing the normalized Laplacian eigenmaps of the affinity matrix
$B$ for various $K$, we see that when $K > 2$, the points appear much
closer to Gaussian mixtures, whereas with $K = 2$, while there is some
separation by community, they do not appear as Gaussian mixtures. 
One possible reason for this may be because we failed to find the
right value for the hyperparameter $\vartheta$ for SSC-ASE. 
}  
\textcolor{blue}{
More specifically, under ideal circumstances, we can set the hyperparameter $\vartheta$ to
the theoretical values that ensure, with high probability, the rows of $\hat{V}$ satisfy the subspace
detection property (SDP). However, as other researchers had pointed
out that, even in the context of sparse subspace clustering where the points lie
exactly on a union of $K$ subspaces, the fact that SDP is satisfied
does not guarantee that the affinity matrix $B$ represents a graph
with $K$ connected components but rather a graph with {\em at least}
$K$ connected component (see e.g., \citep{liu_ssc,sdp_sufficiency,}).
Indeed, we found in our simulations that
setting $\vartheta$ to the theoretical values that guarantees 
SDP often results in $B$ having more than $K$ connected components.
This is why we choose to do spectral clustering on $B$, i.e., we first
compute the normalized Laplacian embeddings using $B$ before
clustering the rows of this embedding using Gaussian mixture models. 
}  
\textcolor{blue}{
Finally we emphasize that in our simulations the proportion of mislabeled
vertices does decreases as $n$ increases. For instance, in section 4.1, if $K = 2$ then the average number of mis-clustered vertices for SSC-ASE is around 20 when $n = 128$ and around 100 when
$n = 4096$; this correspond to an error rate of around $16\%$ and
$2\%$, respectively. In other words SSC-ASE also works quite well for $K =
2$ as the proportion of mis-clustered vertices decreases to $0$ as $n$
increases; it is only when we compared the number of mis-clustered
vertices that $K = 2$ performs worse compared to $K \geq 3$. 
}
